Reasonably Programmable Literal Notation

### 2018-06-25 icfp18/syntax-icfp18.tex
second phase submission

Let us turn now to the perspective of the TLM provider, whose principal task is to define the lexer and parser named in the TLM definition---\li{RegexLexer} and \li{RegexParser} in Fig.~\ref{fig:regex-tlm-def}. Our implementation of TLMs for Reason, called \li{Relit}, assumes that the provider is using \li{ocamllex} \cite{ocaml-manual} and Menhir  \cite{pottier2016menhir} to generate the lexer and parser, respectively. Menhir is a modern derivative of Yacc~\cite{johnson1975yacc,TarditiDR:mly} with support for LR(1) grammars \cite{jourdan2012validating}. The Reason implementation also uses these tools. \li{Relit} requires that the TLM provider package the lexer and parser into an \li{ocamlfind} package named in the TLM definition---in the case of Fig.~\ref{fig:regex-tlm-def}, \li{regex_parser}---so that \li{Relit} can load them, together with their dependencies, at expansion-time. The TLM must also name the starting non-terminal---here, \li{start} on Line~4 of Fig.~\ref{fig:regex-tlm-def}. 
% What is fundamentally important is that there is some clear protocol by which the lexer and parser names and the starting non-terminal name specified by the TLM definition, in this case \li{RegexLexer} and \li{RegexParser.start} on Lines 3-4 of Fig.~\ref{fig:regex-tlm-def}, are resolved to lexer and parser implementations that can be evaluated at expansion-time, i.e. during compilation. 
The \li{ocamllex} and Menhir definitions that implement \li{RegexLexer} and \li{RegexParser} appear in Fig.~\ref{fig:lexer-and-parser}. 
After detailing this example in Sec.~\ref{sec:regex-lexer}-\ref{sec:regexparser}, we summarize the relevant parser correctness criteria in Sec.~\ref{sec:correctness-criteria}, describe how providers might use MetaOCaml to reason more precisely about type correctness in Sec.~\ref{sec:metaocaml}, and finally outline an alternative implementation strategy in Sec.~\ref{sec:html-example} that sidesteps \li{ocamllex} and Menhir to support the use of arbitrary parse functions. In particular, we implement the \li{$html} TLM from Fig.~\ref{fig:nesting-example} using an off-the-shelf HTML parser.

@@

In order to repurpose this existing parse tree representation for proto-expansions, we must provide some way to unambiguously represent splice references, notated earlier in the paper as \li{spliced<m..n : ty>} where \li{m..n} specifies a segment and \li{ty} is the segment's expected type. The \li{ProtoExpr} module solves this problem by providing a function \li{spSURLEURLliced} that takes a segment, of type \li{Segment.t}, and a type representation, of type \li{Parsetree.core_type}, which also comes from OCaml's compiler library, and produces a corresponding value of type \li{ProtoExpr.t} that uniquely represents the corresponding splice reference (see Sec.~\ref{sec:implementation} for more on the internal representation).\todo{say something about splice references in sec 6}{}

@@

Fig.~\ref{fig:html-parser} shows much of the implementation of \li{HtmlParser},  which is constructed by applying a functor that constructs a module that has the same essential interface as the modules that Menhir generates, with the functions \li{expr : string -> ProtoExpr.t} and \li{pat : string -> ProtoPat.t} of the input module exported as the ``non-terminals''. The companion lexer is \li{Relit.TrivLexer}, which simply passes the whole literal body on as a single token. The implementation of \li{expr} on Line 23 of Fig.~\ref{fig:html-parser} reveals that the literal body is parsed by first calling \li{ProtoHtml.parse}, then passing the result on to \li{ExprExpander.expand_proto}. (We discuss \li{pat} in the next section\todo{do this}{}.)

@@

The \li{ExprExpander.expand_proto} function defined, with some helper functions, on Lines~12-21 finishes the job of the TLM parser by mapping the parsed proto-HTML to a proto-expression using the \li{$proto_expr} TLM previously described (in this case, using the \li{$list} antiquotation form that produces a parse tree of OCaml's built in list literal form from a list of parse trees).% It would be straightforward to reuse the \li{ProtoHtml} module for different variations on HTML encodings, implementing only the expander anew. Overall, the amount of code needed to implement HTML literals.

Reason currently builds in ``JSX'' literals, which also support an HTML-like notation with support for splicing in a somewhat idiosynchratic manner motivated by certain JavaScript libraries (in particular, React). The expansion of this notation must be interpretted by a PPX rewriter. The problem is that this makes it difficult to use different HTML-related libraries at once. For example, ReasonReact and an XML library might both want to use this notation. This approach solves this problem (as well as other problems with the PPX-based approach as discussed in Sec.~\ref{sec:existing-approaches}).

@@

@@

For example, the definitions outlined in Fig. \ref{fig:lam} allow us to elegantly express an evaluator for the lambda calculus as shown in Fig. \ref{fig:lam-eval}. We assume that the terms of \li{Lambda} are represented differently from the values to avoid needing additional ``impossible'' cases, but we can use the same notation for both without conflict by defining separate TLMs, \li{$term} and \li{$v}, respectively. Each of these TLMs defines both expression and pattern literal notation, distinguished by qualifiers in the TLM definition as shown on Lines 5-6 of Fig. \ref{fig:lam}, and we see examples of all four of these notations in use in Fig. \ref{fig:lam-eval}. These TLMs were designed to be used for operating on \li{Lambda} terms, so we chose a splicing convention inspired by the typical convention ``on paper'', where identifiers appearing in the literal are parsed as spliced (OCaml) variables, as indicated by the colors in Fig. \ref{fig:lam-eval}. For example, on Line 5, \li{x} and \li{e} are bound by the literal pattern for \li{Lambda} terms on the left, then spliced into the expression literal on the right to construct the corresponding \li{Lambda} value. % A different pair of TLMs at the same types could make the opposite choice, parsing identifiers as \li{Lambda} language variables and defining a separate notation for splicing. %These would be less elegant for writing the evaluator but more useful when, for example, writing its tests. This sort of flexibility is difficult to achieve in settings where the grammar of the host language is itself extended (e.g. with the more limited mixfix systems of Coq \cite{Coq:manual} or Agda \cite{norell2007towards}) because syntactic conflicts ensue.

From a client programmer's perspective, reasoning principles analogous to those described in Sec. \ref{sec:client-perspective} for expression TLMs are available. \textbf{Responsibility} is assigned by the same protocol, and the type annotation on the responsible TLM governs the type of the generated pattern, satisfying the \textbf{Expansion Typing} condition. Patterns can contain module paths and type annotations, so pattern TLMs are governed by the same \textbf{Context Independence} condition as expression TLMs. The same protocol around \textbf{Segmentation} is also enforced.

\begin{figure}[t]
\begin{subfigure}[t]{0.55\textwidth}
\begin{lstlisting}[morekeywords={expression}]
  module Lambda = {/* a typical encoding */};
  module LambdaNotation {
    notation $term at Lambda.term { 
      lexer SimpleLangLexer;
      expression parser LambdaParser.term_e;
      pattern parser LambdaParser.term_p;
      dependencies 
        { module Lambda = Lambda } };
    notation $v at Lambda.v {/* analogous */}
  };
\end{lstlisting}
\vspace{-6px}
\caption{Expression and pattern TLMs for \lismall{Lambda} terms and values}
\label{fig:lam}
\vspace{2px}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.42\textwidth}
\begin{lstlisting}
  open LambdaNotation;
  exception Unbound(Lambda.var);
  let rec eval = $term.(fun 
  | `(x)` => raise Unbound(x)
  | `(SURLlamEURL xSURL.EURLe)` => $v `(SURLlamEURL xSURL.EURLe)`
  | `(e1SURL(EURLe2SURL)EURL)` => {
    let $v `(SURLlamEURL xSURL.EURLe)` = eval(e1);
    let v2 = eval(e2);
    eval(`(SURL[EURLv2SURL/EURLxSURL]EURLe)`) 
  });
\end{lstlisting}
\vspace{-6px}
\caption{A reasonably elegant \lismall{Lambda} evaluator}
\label{fig:lam-eval}
\end{subfigure}
\vspace{2px}
\caption{Expression and pattern literal notation for lambda terms and values}
\vspace{-8px}
\label{fig:lam-example}
\end{figure}

The main novelty has to do with \textbf{Segment Typing} and \textbf{Capture}. Variables in patterns do not refer to existing bindings, as in expressions, but rather introduce bindings into other expressions, e.g. the corresponding branch of a case analysis or the body of a function, so the critical question is this: given only the segmentation of a pattern literal, can we determine exactly which variables the expansion binds, and what the types of these variables are? To answer in the affirmative, we need to preclude the possibility of ``invisible bindings'', so the system ensures that pattern literals bind only those variables that appear inside spliced patterns. This implies that \emph{proto-patterns simply cannot contain pattern variables}. The segmentation assigns a type to each spliced segment, so the question just asked can be answered given only the segmentation.

From the provider's perspective, the parser specified by the \li{pattern parser} clause must generate a proto-pattern, rather than a proto-expression. In Reason, this is a value of type \li{Relit.ProtoPat.t}. Much as with \li{Relit.ProtoExpr.t} from Fig. \ref{fig:relit-util}, this type is defined as a synonym for \li{Parsetree.patteSURLEURLrn} from OCaml's compiler library equipped with an additional function, \li{Relit.ProtoPat.spSURLEURLliced}, that constructs splice references. In principle, it might be better to define \li{Relit.ProtoPat.t} to exclude the representation of pattern variables, and we do just that in the theoretical development in Sec. \ref{sec:setlms-formally}, but \li{Relit} checks this straightforward condition during proto-expansion validation. Note that in OCaml, boolean guards can be associated with rules of a \li{match} (in Reason, \li{switch}) expression, but the guard is not part of the syntax of patterns. (If it were, we would need to validate it as in Sec.~\ref{sec:setlms}). %In order to ensure that a pattern TLM can be applied anywhere a pattern can appear, e.g. in a \li{let} expression as on Line 7 of Fig. \ref{fig:lam-eval}, a pattern TLM cannot generate a guard. % In a language where boolean guards are included in the syntax of patterns, we would need to enforce context independence, capture avoidance and segmentation-related conditions on guard expressions as well.%  In languages where boolean guards can appear at any depth in a pattern, the same

@@

Fig. \ref{fig:map-sig} shows a portion of the signature \li{Map.S} from the OCaml standard library, which specifies a polymorphic abstract data type \cite{liskov1974programming,harper1997programming} \li{t('a)} of finite maps with keys of type \li{key}. There are many ways to implement this signature. For example, the \li{Map.Make} functor in the standard library implements this signature using balanced binary trees given a module that specifies a \li{key} type equipped with comparison functions. The \li{Misc.StringMap} module in the standard library also provides a specialized implementation of this signature with type \li{key = string}.

We can define a TLM, \li{$map}, that is parametric over implementations of this signature, \li{M : Map.S}, and over choices of the co-domain type, \li{a}, as shown in Fig. \ref{fig:map-tlm}. We can then partially apply \li{$map} to \li{Misc.StringMap} as shown on Line 1 of Fig.~\ref{fig:map-ap}, producing a TLM, \li{$stringmap},  parameterized only over the type \li{a}, with expansion type \li{Misc.StringMap.t(a)}. We can apply \li{$stringmap} to a type specialize it further, e.g. \li{$stringmap string} has expansion type \li{Misc.StringMap.t(string)}.  For TLMs where all remaining parameters are types mentioned in the expansion type, we can also immediately apply the TLM to a generalized literal form as shown on Lines~2-3 of Fig.~\ref{fig:map-ap}. The type parameters are determined by unification using the types inferred for the spliced expressions together with the segment types specified in the segmentation, and any surrounding type constraints. In this example, we can infer \li{a} to be \li{string} because \li{string_of_int(id) : string} and the corresponding segment type is \li{a}. TLM abbreviations can themselves be parameterized to support partial application of parameters other than the last.

@@

\begin{figure}
\begin{subfigure}[t]{0.4\textwidth}
\begin{lstlisting}
  module Map = { 
    module type S = {
      type key;
      type t('a);
      let empty : t('a);
      let add : 
        key => 'a => t('a) => t('a);
    /* ... */ }
  }
\end{lstlisting}
\vspace{-7px}
\caption{The \li{Map.S} signature}
\label{fig:map-sig}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.55\textwidth}
\begin{subfigure}[t]{\textwidth}
\begin{lstlisting}
  notation $map(M : Map.S, type a) at M.t(a) {
    lexer  MapLexer; parser MapParser; 
  }
\end{lstlisting}
\vspace{-6px}
\caption{The parametric TLM \li{$map}}
\vspace{5px}
\label{fig:map-tlm}
\end{subfigure}
\begin{subfigure}[b]{\textwidth}
\begin{lstlisting}
  notation $stringmap = $map(Misc.StringMap);
  let make_request id req v = $stringmap
    `( "session_id" SURL->EURL string_of_int(id)SURL,EURL
                req SURL->EURL v )`;
\end{lstlisting}
\vspace{-4px}
\caption{Applying \li{$map}}
\label{fig:map-ap}
\end{subfigure}
\end{subfigure}
\vspace{4px}
\caption{Literal notation for all finite map implementations using parametric TLMs.}
\vspace{-5px}
\label{fig:map-fig}
\end{figure}

@@

An alternative point-of-view in ML is to treat the \li{dependencies} clause as fundamental and achieve parameterization using module functions (functors), e.g. we could express \li{$map} as follows:
\begin{lstlisting}[numbers=none]
  module MapNotation = (M : Map.S, A : { type a; }) => {
    notation $map at M.t(A.a) { 
      lexer MapLexer; parser MapParser; 
      dependencies { module M = M; type a = A.a; } } }
\end{lstlisting}
% The conveniences related to type parameter inference are not available when parameterization is encoded in this way.

@@

This section will present a calculus of simple expression and pattern TLMs based on ``core ML'' called $\miniVersePat$. By the end of this section, we will have a theorem that formalizes the six reasoning principles that were outlined informally in the previous sections. $\miniVersePat$ consists of an \emph{unexpanded language}, or \emph{UL}, defined by typed expansion to the core language, named for the purposes of this paper $\miniVerseBase$.

@@

\subsection{Syntax of the Unexpanded Language}\label{sec:syntax-U}\label{sec:s-UL}
Fig. \ref{fig:U-unexpanded-terms} defines the syntax of the unexpanded language. Unlike core language types and expressions, unexpanded expressiona and types are \textbf{not} abstract binding trees -- we do \textbf{not} assume the standard notions of renaming, alpha-equivalence or substitution. Instead, they are simple inductive  structures. This is because unexpanded expressions remain ``partially parsed'' due to the presence of literal bodies, $b$, from which spliced terms might be extracted during expansion. In fact, unexpanded types and expressions do not involve variables at all, but rather \emph{identifiers}, $\ut$ and $\ux$. % This distinction between identifiers and variables is technically crucial to our developments. %We \textbf{cannot} adopt the usual definitions of $\alpha$-renaming of identifiers, because unexpanded types and expressions are still in a ``partially parsed'' state -- the literal bodies, $b$, within an unexpanded expression might contain spliced subterms that are ``surfaced'' by a TLM only during typed expansion, as we will detail below. %identifiers are given meaning by expansion to variables. %In other words, unexpanded expressions are not abstract binding trees, nor sequences of characters, but a ``transitional'' structure with some characteristics of each of these. 
%For this reason, we will need to handle generating fresh variables explicitly at binding sites in our semantics. %To do so, we distinguish \emph{type identifiers}, $\ut$, and \emph{expression identifiers}, $\ux$, from type variables, $t$, and expression variables, $x$. identifiers will be given meaning by expansion to variables (which, in turn, are given meaning by substitution, as described above).

@@

Most of the unexpanded forms in Figure \ref{fig:U-unexpanded-terms}  mirror the expanded forms. We refer to these as the \emph{common forms}. % The mapping from expanded forms to common unexpanded forms is defined explicitly in the accompanying technical report. 
The typed expansion rules that handle common forms mirror the corresponding typing rules. The \emph{expression TLM context}, $\uPsi$, and the \emph{pattern TLM context}, $\uPhi$, detailed below pass through these rules opaquely. For example, the rules for variables and lambdas are shown being applied in the example derivation in Fig.~\ref{fig:expansion-exmpl}, discussed below. The full set of rules is in the accompanying technical report.
%Each of these rules is based on the corresponding typing rule, i.e. Rules (\ref{rule:hastypeU-var}) through (\ref{rule:hastypeU-case}), respectively. For example, the following typed expansion rules are based on the typing rules (\ref{rule:hastypeU-var}), (\ref{rule:hastypeU-lam}) and (\ref{rule:hastypeU-ap}), respectively:% for unexpanded expressions of variable, function and application form, respectively: 
% {\small\begin{mathpar}
%   \inferrule[ee-id]{ }{\expandsUP{\uDelta}{\uGamma, \uGhyp{\ux}{x}{\tau}}{\uPsi}{\uPhi}{\ux}{x}{\tau}}

@@

The type abbreviated $\tCEExp$ classifies encodings of \emph{proto-expressions}, $\ce$ (pronounced ``grave $e$''). The syntax of proto-expressions, defined in Fig.~\ref{fig:U-candidate-terms}, will be described when we describe proto-expansion validation in Sec. \ref{sec:ce-syntax-U}. 
As with $\tBody$, we need only take as a condition that there is an isomorphism between values of type $\tCEExp$ and closed proto-expressions, which is mediated in one direction by the \emph{proto-expression decoding judgement}, $\decodeCondE{e}{\ce}$ that we will return to when we describe TLM application below (see accompanying technical report for the full condition).

@@

The fourth premise applies the parse function $\eparse$ to the encoding of the literal body. If parsing succeeds, i.e. a value of the form $\aein{\mathtt{SuccessE}}{\ecand}$ results from evaluation, then $\ecand$ will be a value of type $\tCEExp$ (by type safety and the canonical forms lemma, assuming a well-formed expression TLM context). We call $\ecand$ the \emph{encoding of the proto-expansion}. If the parse function produces a value labeled $\lbltxt{Error}$, then typed expansion fails and formally, no rule is necessary.

>>>>

Let us turn now to the perspective of the TLM provider, whose principal task is to define the lexer and parser named in the TLM definition---\li{RegexLexer} and \li{RegexParser} in Fig.~\ref{fig:regex-tlm-def}. Our implementation of TLMs for Reason, called \li{Relit}, assumes that the provider is using \li{ocamllex} \cite{ocaml-manual} and Menhir  \cite{pottier2016menhir,DBLP:journals/entcs/PottierR06} to generate the lexer and parser, respectively. Menhir is a modern derivative of Yacc~\cite{johnson1975yacc} with support for LR(1) grammars \cite{jourdan2012validating,DBLP:journals/entcs/PottierR06}. The Reason implementation also uses these tools. \li{Relit} requires that the TLM provider package the lexer and parser into an \li{ocamlfind} package named in the TLM definition---in Fig.~\ref{fig:regex-tlm-def}, \li{regex_parser}---so that \li{Relit} can load them, together with their dependencies, at expansion-time. The TLM must also name the starting non-terminal---here, \li{start} on Line~4 of Fig.~\ref{fig:regex-tlm-def}. 
% What is fundamentally important is that there is some clear protocol by which the lexer and parser names and the starting non-terminal name specified by the TLM definition, in this case \li{RegexLexer} and \li{RegexParser.start} on Lines 3-4 of Fig.~\ref{fig:regex-tlm-def}, are resolved to lexer and parser implementations that can be evaluated at expansion-time, i.e. during compilation. 
The \li{ocamllex} and Menhir definitions that implement \li{RegexLexer} and \li{RegexParser} appear in Fig.~\ref{fig:lexer-and-parser}. 
After detailing this example in Sec.~\ref{sec:regex-lexer}-\ref{sec:regexparser}, we summarize the relevant correctness criteria in Sec.~\ref{sec:correctness-criteria}, describe how providers might use MetaOCaml to reason more precisely about type correctness in Sec.~\ref{sec:metaocaml}, and outline an alternative implementation strategy in Sec.~\ref{sec:html-example} that sidesteps \li{ocamllex} and Menhir to support the use of arbitrary parse functions. In particular, we implement the \li{$html} TLM from Fig.~\ref{fig:nesting-example} using an off-the-shelf HTML parser.

@@

In order to repurpose this existing parse tree representation for proto-expansions, we must provide some way to unambiguously represent splice references, notated earlier in the paper as \li{spliced<m..n : ty>} where \li{m..n} specifies a segment and \li{ty} is the segment's expected type. The \li{ProtoExpr} module solves this problem by providing a function \li{spSURLEURLliced} that takes a segment, of type \li{Segment.t}, and a type representation, of type \li{Parsetree.core_type}, which also comes from OCaml's compiler library, and produces a corresponding value of type \li{ProtoExpr.t} that uniquely represents the corresponding splice reference (see Sec.~\ref{sec:implementation} for more on the internal representation).

@@

Fig.~\ref{fig:html-parser} shows much of the implementation of \li{HtmlParser},  which is constructed by applying a functor that constructs a module that has the same essential interface as the modules that Menhir generates, with the functions \li{expr : string -> ProtoExpr.t} and \li{pat : string -> ProtoPat.t} of the input module exported as the ``non-terminals''. The companion lexer is \li{Relit.TrivLexer}, which simply passes the whole literal body on as a single token. The implementation of \li{expr} on Line 23 of Fig.~\ref{fig:html-parser} reveals that the literal body is parsed by first calling \li{ProtoHtml.parse}, then passing the result on to \li{ExprExpander.expand_proto}. (We discuss \li{pat} in the next section.)

@@

The \li{ExprExpander.expand_proto} function defined, together with helper functions, on Lines~12-21 finishes the job of the TLM parser by mapping the parsed proto-HTML to a proto-expression using the \li{$proto_expr} TLM previously described (in this case, using the \li{$list} antiquotation form that produces a parse tree of OCaml's built in list literal form from a list of parse trees).% It would be straightforward to reuse the \li{ProtoHtml} module for different variations on HTML encodings, implementing only the expander anew. Overall, the amount of code needed to implement HTML literals.

Reason currently builds in ``JSX'' literals, which also support an HTML-like notation with support for splicing. The expansion of this notation must be interpretted by a PPX rewriter. The problem is that this makes it difficult to use different HTML-related libraries at once. For example, ReasonReact and an XML library might both want to use this notation. This approach solves this problem (as well as other problems with the PPX-based approach as discussed in Sec.~\ref{sec:existing-approaches}).

@@

\begin{figure}[t]
\begin{subfigure}[t]{0.55\textwidth}
\begin{lstlisting}[morekeywords={expression}]
  module Lambda = {/* a typical encoding */};
  module LambdaNotation = {
    notation $term at Lambda.term { 
      lexer LambdaLexer
      expression parser LambdaParser.term_e
      pattern parser LambdaParser.term_p
      in package lambda_parser;
      dependencies = {module Lambda=Lambda} };
    notation $v at Lambda.v {/* analogous */}
  };
\end{lstlisting}
\vspace{-6px}
\caption{Expression and pattern TLMs for \lismall{Lambda} terms and values}
\label{fig:lam}
\vspace{2px}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.42\textwidth}
\begin{lstlisting}
  open LambdaNotation;
  exception Unbound(Lambda.var);
  let rec eval = $term.(fun 
  | `(x)` => raise Unbound(x)
  | `(SURLlamEURL xSURL.EURLe)` => $v `(SURLlamEURL xSURL.EURLe)`
  | `(e1SURL(EURLe2SURL)EURL)` => {
    let $v `(SURLlamEURL xSURL.EURLe)` = eval(e1);
    let v2 = eval(e2);
    eval(`(SURL[EURLv2SURL/EURLxSURL]EURLe)`) 
  });
\end{lstlisting}
\vspace{-6px}
\caption{A reasonably elegant \lismall{Lambda} evaluator}
\label{fig:lam-eval}
\end{subfigure}
\caption{Expression and pattern literal notation for lambda terms and values}
\vspace{-8px}
\label{fig:lam-example}
\end{figure}

@@

For example, the definitions outlined in Fig. \ref{fig:lam} allow us to elegantly express an evaluator for the lambda calculus as shown in Fig. \ref{fig:lam-eval}. We assume that the terms of \li{Lambda} are of type \li{Lambda.term}, and the values have a different representation, of type \li{Lambda.v}, to avoid needing additional ``impossible'' cases. We can provide common notation for both without conflict by defining separate TLMs, \li{$term} and \li{$v}, respectively, in Fig.~\ref{fig:lam}. Each of these TLMs defines both expression and pattern literal notation, distinguished by qualifiers in the TLM definition as shown on Lines 5-6 of Fig. \ref{fig:lam}, and we see examples of all four of these notations in use in Fig. \ref{fig:lam-eval}. These TLMs were designed to be used for operating on \li{Lambda} terms, so we chose a splicing convention inspired by the typical convention ``on paper'', where identifiers appearing in the literal are parsed as spliced (OCaml) variables, as indicated by the colors in Fig. \ref{fig:lam-eval}. % For example, on Line 5, \li{x} and \li{e} are bound by the literal pattern for \li{Lambda} terms on the left, then spliced into the expression literal on the right. % A different pair of TLMs at the same types could make the opposite choice, parsing identifiers as \li{Lambda} language variables and defining a separate notation for splicing. %These would be less elegant for writing the evaluator but more useful when, for example, writing its tests. This sort of flexibility is difficult to achieve in settings where the grammar of the host language is itself extended (e.g. with the more limited mixfix systems of Coq \cite{Coq:manual} or Agda \cite{norell2007towards}) because syntactic conflicts ensue.

\subsection{Client Perspective} From a client programmer's perspective, reasoning principles analogous to those described in Sec.~\ref{sec:client-perspective} for expression TLMs are available. \textbf{Responsibility} is assigned by the same protocol, and the type annotation on the responsible TLM governs the type of the generated pattern, satisfying the \textbf{Expansion Typing} condition. Patterns can contain module paths and type annotations, so pattern TLMs are governed by the same \textbf{Context Independence} condition as expression TLMs. The same protocol around \textbf{Segmentation} is also enforced. The segmentation assigns a type to each spliced segment, so we can also reason about \textbf{Segment Typing}.

Variables in patterns do not refer to existing bindings, as in expressions, so we do not need to worry about capture avoidance. Instead, patterns introduce bindings into other expressions, e.g. the corresponding branch of a case analysis, so the critical question is this: given only the segmentation of a pattern literal, how can we determine exactly which variables the expansion binds? To answer, we need to preclude the possibility of ``invisible bindings'', so the system ensures that pattern literals bind only those variables that appear inside spliced patterns. We call this property \textbf{Visibility}.

Note that in OCaml, boolean guards can be associated with rules of a \li{match} expression, but the guard is not part of the syntax of patterns. (If it were, we would need to validate it as in Sec.~\ref{sec:setlms}.) %In order to ensure that a pattern TLM can be applied anywhere a pattern can appear, e.g. in a \li{let} expression as on Line 7 of Fig. \ref{fig:lam-eval}, a pattern TLM cannot generate a guard. % In a language where boolean guards are included in the syntax of patterns, we would need to enforce context independence, capture avoidance and segmentation-related conditions on guard expressions as well.%  In languages where boolean guards can appear at any depth in a pattern, the same

\subsection{Provider Perspective}
From the provider's perspective, the parser specified by the \li{pattern parser} clause differs only in that it must generate a proto-pattern, rather than a proto-expression. In Reason, this is a value of type \li{Relit.ProtoPat.t}. Much as with \li{Relit.ProtoExpr.t} from Fig. \ref{fig:relit-util}, this type is defined as a synonym for \li{Parsetree.patteSURLEURLrn} from OCaml's compiler library equipped with an additional function, \li{Relit.ProtoPat.spSURLEURLliced}, that constructs splice references. The \textbf{Visibility} property is maintained by enforcing the rule that a \emph{proto-pattern simply cannot contain pattern variables}.

We omit the straightforward details of the parsers in Fig.~\ref{fig:lam-example}. In cases where the output of an expression parser uses only constructors and constants, like the HTML example in Fig.~\ref{fig:html-parser}, we can simply call the expression parser and then convert the result to a proto-pattern (Line 24 of Fig.~\ref{fig:html-parser}). %In cases where an expression parser uses other constructs, e.g. the \li{let} bindings used in the regex parser in Fig.~\ref{fig:regex-parser}, we would need to define it separately.

 % In principle, it might be better to define \li{Relit.ProtoPat.t} to exclude the representation of pattern variables, and we do just that in the theoretical development in Sec. \ref{sec:setlms-formally}, but \li{Relit} checks this straightforward condition during proto-expansion validation.

@@

Fig. \ref{fig:map-sig} shows a portion of the signature \li{Map.S} from the OCaml standard library, which specifies a polymorphic abstract data type \li{t('a)} of finite maps with keys of type \li{key}. There are many ways to implement this signature. For example, the \li{Map.Make} functor in the standard library implements this signature using balanced binary trees given a module that specifies a \li{key} type equipped with comparison functions. The \li{Misc.StringMap} module in the standard library also provides a specialized implementation of this signature with type \li{key = string}.

We can define a TLM, \li{$map}, that is parametric over implementations of this signature, \li{M : Map.S}, and over choices of the co-domain type, \li{a}, as shown in Fig. \ref{fig:map-tlm}. We can then partially apply \li{$map} to \li{Misc.StringMap} as shown on Line 1 of Fig.~\ref{fig:map-ap}, producing a TLM, \li{$stringmap},  parameterized only over the type \li{a}, with expansion type \li{Misc.StringMap.t(a)}. We might apply \li{$stringmap} to a type specialize it further, e.g. \li{$stringmap string} has expansion type \li{Misc.StringMap.t(string)}. Alternatively, for TLMs where all remaining parameters are types mentioned in the expansion type, we can immediately apply the TLM to a generalized literal form as shown on Lines~2-3 of Fig.~\ref{fig:map-ap}. The type parameters are determined by unification using the types inferred for the spliced expressions together with the segment types specified in the segmentation, and any surrounding type constraints. In this example, we can infer \li{a} to be \li{string} because \li{string_of_int(id) : string} and the corresponding segment type is \li{a}. TLM abbreviations can themselves be parameterized to support partial application of parameters other than the last.

@@

\begin{figure}
\begin{subfigure}[t]{0.4\textwidth}
\begin{lstlisting}
  module Map = { 
    module type S = {
      type key;
      type t('a);
      let empty : t('a);
      let add : 
        key => 'a => t('a) => t('a);
      /* ... */
    };
  };
\end{lstlisting}
\vspace{-7px}
\caption{The \li{Map.S} signature}
\label{fig:map-sig}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.55\textwidth}
\begin{subfigure}[t]{\textwidth}
\begin{lstlisting}
  notation $map(M : Map.S, type a) at M.t(a) {
    lexer MapLexer parser MapParser 
    in package map_parser;
  }
\end{lstlisting}
\vspace{-8px}
\caption{The parametric TLM \li{$map}}
\vspace{5px}
\label{fig:map-tlm}
\end{subfigure}
\begin{subfigure}[b]{\textwidth}
\begin{lstlisting}
  notation $stringmap = $map(Misc.StringMap);
  let make_request id req v = $stringmap
    `( "session_id" SURL->EURL string_of_int(id)SURL,EURL
                req SURL->EURL v )`;
\end{lstlisting}
\vspace{-4px}
\caption{Applying \li{$map}}
\label{fig:map-ap}
\end{subfigure}
\end{subfigure}
\vspace{4px}
\caption{Literal notation for all finite map implementations using parametric TLMs.}
\vspace{-5px}
\label{fig:map-fig}
\end{figure}

@@

An alternative point-of-view in ML is to treat the \li{dependencies} clause as fundamental and achieve parameterization using module functions (functors), e.g. we could express \li{$map} as follows:
\begin{lstlisting}[numbers=none]
  module MapNotation = (M : Map.S, A : { type a; }) => {
    notation $map at M.t(A.a) { 
      lexer MapLexer parser MapParser in package map_parser;
      dependencies { module M = M; type a = A.a; } } }
\end{lstlisting}
% The conveniences related to type parameter inference are not available when parameterization is encoded in this way.

@@

This section presents  $\miniVersePat$, a calculus of simple expression and pattern TLMs based on ``core ML''. By the end of this section, we will have theorems that formalize the reasoning principles that were outlined informally in the previous sections. $\miniVersePat$ consists of an {unexpanded language} ({UL}) defined by typed expansion to the core language, named for the purposes of this paper $\miniVerseBase$.

@@

\subsection{Syntax of the Unexpanded Language}\label{sec:syntax-U}\label{sec:s-UL}
Fig. \ref{fig:U-unexpanded-terms} defines the syntax of the unexpanded language. Unlike core language types and expressions, unexpanded expressions and types are \textbf{not} abstract binding trees -- we do \textbf{not} assume the standard notions of renaming, alpha-equivalence or substitution. Instead, they are simple inductive  structures. This is because unexpanded expressions remain ``partially parsed'' due to the presence of literal bodies, $b$, from which spliced terms might be extracted during expansion. In fact, unexpanded types and expressions do not involve variables at all, but rather \emph{identifiers}, $\ut$ and $\ux$. % This distinction between identifiers and variables is technically crucial to our developments. %We \textbf{cannot} adopt the usual definitions of $\alpha$-renaming of identifiers, because unexpanded types and expressions are still in a ``partially parsed'' state -- the literal bodies, $b$, within an unexpanded expression might contain spliced subterms that are ``surfaced'' by a TLM only during typed expansion, as we will detail below. %identifiers are given meaning by expansion to variables. %In other words, unexpanded expressions are not abstract binding trees, nor sequences of characters, but a ``transitional'' structure with some characteristics of each of these. 
%For this reason, we will need to handle generating fresh variables explicitly at binding sites in our semantics. %To do so, we distinguish \emph{type identifiers}, $\ut$, and \emph{expression identifiers}, $\ux$, from type variables, $t$, and expression variables, $x$. identifiers will be given meaning by expansion to variables (which, in turn, are given meaning by substitution, as described above).

@@

Most of the unexpanded forms in Figure \ref{fig:U-unexpanded-terms}  mirror the expanded forms. We refer to these as the \emph{common forms}. % The mapping from expanded forms to common unexpanded forms is defined explicitly in the accompanying technical report. 
The typed expansion rules that handle common forms mirror the corresponding typing rules. The \emph{expression TLM context}, $\uPsi$, and the \emph{pattern TLM context}, $\uPhi$, detailed below pass through these rules opaquely. For example, the rules for variables and lambdas are shown being applied in the example derivation in Fig.~\ref{fig:expansion-exmpl}, discussed below. The full set of rules is in the accompanying technical report \cite{icfp18tr}.
%Each of these rules is based on the corresponding typing rule, i.e. Rules (\ref{rule:hastypeU-var}) through (\ref{rule:hastypeU-case}), respectively. For example, the following typed expansion rules are based on the typing rules (\ref{rule:hastypeU-var}), (\ref{rule:hastypeU-lam}) and (\ref{rule:hastypeU-ap}), respectively:% for unexpanded expressions of variable, function and application form, respectively: 
% {\small\begin{mathpar}
%   \inferrule[ee-id]{ }{\expandsUP{\uDelta}{\uGamma, \uGhyp{\ux}{x}{\tau}}{\uPsi}{\uPhi}{\ux}{x}{\tau}}

@@

The type abbreviated $\tCEExp$ classifies encodings of \emph{proto-expressions}, $\ce$ (pronounced ``grave $e$''). The syntax of proto-expressions, defined in Fig.~\ref{fig:U-candidate-terms}, will be described when we describe proto-expansion validation in Sec. \ref{sec:ce-syntax-U}. 
As with $\tBody$, we need only take as a condition that there is an isomorphism between values of type $\tCEExp$ and closed proto-expressions, which is mediated in one direction by the \emph{proto-expression decoding judgement}, $\decodeCondE{e}{\ce}$ that we will return to when we describe TLM application below (see accompanying technical report for the full conditions).

@@

The fourth premise applies the parse function $\eparse$ to the encoding of the literal body. If parsing succeeds, i.e. a value of the form $\aein{\mathtt{SuccessE}}{\ecand}$ results from evaluation, then $\ecand$ will be a value of type $\tCEExp$ (by type safety). We call $\ecand$ the \emph{encoding of the proto-expansion}. If the parse function produces a value labeled $\lbltxt{Error}$, then typed expansion fails and no rule is necessary.
